Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-07-17 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
Last Revision: small changes to improve readibility

%, leading inefficient flow sweeping which can bypass significant quantities of recoverable oil and may lead to early breakthrough of water into neighbour production wells.
%In hydrocarbon reservoir exploration, viscous %and density 
%instabilities are relatively common in waterflooding processes. %As water and oil interacts, the interface between these fluids moves creating an uneven or fingered flow profile (\ie collapse of the interface between fluids, Fig.~\ref{fig:simple_case}). 
%Viscous instabilities result in inefficient flow sweeping which can bypass significant quantities of recoverable oil and may lead to early breakthrough of water into neighbour production wells. %Viscous instabilities are mainly controlled by the mobility ratio (MR) between displacing and displaced fluids. Other conditions that may also influence the severity of viscous fingering are: heterogeneity (\ie wide spatial porosity and/or permeability distribution), gravitational forces, anisotropic dispersion, non-monotonic viscosity profile etc \citep{budek_2017,nicolaides_2015}.
 Under the assumption that fluids remain immiscible along the interface, surface tension plays an important role in determining shape and progress of the fingers~\citep{howison_2000}. During the displacement of a fluid by a less viscous one, the expected uniform front \citep{buckley_1942,sheldon_1959} is perturbed leading to an uneven front with elongations at the outside edge of the fluid interface. %In homogeneous domains, fingers start to develop when the surface tension acting on the interface between the fluids exert an opposite force towards the change of shape of the interface. The interface becomes unstable and collapses, taking a curved shape \citep{homsy_1987, jackson_2017}. 
In immiscible displacements, viscous fingering occurs when the viscosity ratio is greater than unity. As surface tension becomes weak, the interface is stressed and becomes unstable leading to the formation of fingers. %At this point it should be mentioned that there are two parameters -- Peclet number (Pe) and mobility ratio (MR), that determine the flow stability characteristics. Out of these finger shape formation there are always a few dominant fingers that spread and shield the growth of other fingers. 
The interface of the main finger collapses and starts splitting into new lobes of fingers. One of these new fingers may eventually outgrow the others and then spreads to occupy an increasingly larger width. In the process, the finger reaches a critical width while the saturation gradient at the front becomes steep as a result of stretching caused by the cross-flow, causing the tip of the finger to become unstable and splitting again, and the pattern repeats. Therefore, surface tension plays an essential dual role, it must be weak enough for the tip front to be unstable, but it is also the physical force causing the spreading and ensuing repeated branching \citep{tan_1988, carvalho_2013}. In heterogeneous domains, such instability may be triggered by permeability differences across regions as shown in Section~\ref{section:results_homo_hete}. %The higher the velocity of the low viscosity fluid, the less wide (tip-splitting behaviour) the finger is. %Pressure differences acting on the interface produces a net pressure force,      
%\begin{equation} 
%\Delta P= - \gamma \nabla\cdot\hat{n}. 
%\label{eq:pressure_dif} 
%\end{equation}
%\noindent This expression is also know as the Young-Laplace equation, a relation describing the capillary pressure across the interface between two fluids, with $\Delta P$ denoting the pressure difference and $\hat{n}$ is the unit normal vector out of the surface.

@@

The MR expression (Eqn.~\ref{eq:MR}) contains phase saturations, $S_{w}\left(x_{i},t\right)$ and  $S_{nw}\left(x_{i},t\right)$ (Eqns.~\ref{Eqn:CoreyBrooks1}-\ref{Eqn:CoreyBrooks2}), however it is clear that during fluid displacement they change in time and space according to the global mass balance equation. Assuming that phase viscosities remain invariant throughout the simulated fluid displacement, in this work, with no lack of generality, the MR can be replaced by the viscosity ratio (VR),
\begin{displaymath} 
    \text{VR} = \frc{\mu_{i}}{\mu_{j}},
\end{displaymath}
which will be used in the parametrisation of the numerical simulations conducted in Section~\ref{section:results}. Analysis performed in the following sections will make use of this simplified definition as phase saturation $\left(S_{\alpha}\right)$ is a time- and spatial-dependent prognostic field which is calculated along with pressure ($p$) and velocity $\left(\mathbf{u}_{\alpha}\right)$ variables.

\medskip
In the next section, the numerical formulation used to simulate multi-fluid flow in porous media is briefly validated (Section \ref{section:results_initial_model_validation}) against laboratory experiments (qualitative validation). The impact of VR (quantitative validation) and heterogeneity on the onset instability and growth of fingers are numerically investigated in Section \ref{section:results_homo_hete}. Section \ref{section:results_hete_fix_adapt} demonstrates the importance of appropriate mesh resolution to adequately capture the initial stages of viscous fingers formation and development. Finally, flow pathway (channelling) is the focus of Section \ref{section:results_3D}.

>>>>

%, leading inefficient flow sweeping which can bypass significant quantities of recoverable oil and may lead to early breakthrough of water into neighbour production wells.
%In hydrocarbon reservoir exploration, viscous %and density 
%instabilities are relatively common in waterflooding processes. %As water and oil interacts, the interface between these fluids moves creating an uneven or fingered flow profile (\ie collapse of the interface between fluids, Fig.~\ref{fig:simple_case}). 
%Viscous instabilities result in inefficient flow sweeping which can bypass significant quantities of recoverable oil and may lead to early breakthrough of water into neighbour production wells. %Viscous instabilities are mainly controlled by the mobility ratio (MR) between displacing and displaced fluids. Other conditions that may also influence the severity of viscous fingering are: heterogeneity (\ie wide spatial porosity and/or permeability distribution), gravitational forces, anisotropic dispersion, non-monotonic viscosity profile etc \citep{budek_2017,nicolaides_2015}.
 Under the assumption that fluids remain immiscible along the interface, surface tension plays an important role in determining shape and progress of the fingers~\citep{howison_2000}. During the displacement of a fluid by a less viscous one, the expected uniform front \citep{buckley_1942,sheldon_1959} is perturbed leading to an uneven front with elongations at the outside edge of the fluid interface. %In homogeneous domains, fingers start to develop when the surface tension acting on the interface between the fluids exert an opposite force towards the change of shape of the interface. The interface becomes unstable and collapses, taking a curved shape \citep{homsy_1987, jackson_2017}. 
In immiscible displacements, viscous fingering occurs when the viscosity ratio is greater than unity. As surface tension becomes weak, the interface is stressed and becomes unstable leading to the formation of fingers. %At this point it should be mentioned that there are two parameters -- Peclet number (Pe) and mobility ratio (MR), that determine the flow stability characteristics. Out of these finger shape formation there are always a few dominant fingers that spread and shield the growth of other fingers. 
The interface of the main finger collapses and starts splitting into new lobes of fingers. One of these new fingers may eventually outgrow the others and then spreads to occupy an increasingly larger width. In the process, the finger reaches a critical width while the saturation gradient at the front becomes steep as a result of stretching caused by the cross-flow, causing the tip of the finger to become unstable and splitting again, and the pattern repeats.

Therefore, surface tension plays an essential dual role, it must be weak enough for the tip front to be unstable, but it is also the physical force causing the spreading and ensuing repeated branching \citep{tan_1988, carvalho_2013}. In heterogeneous domains, such instability may be triggered by permeability differences across regions as shown in Section~\ref{section:results_homo_hete}. %The higher the velocity of the low viscosity fluid, the less wide (tip-splitting behaviour) the finger is. %Pressure differences acting on the interface produces a net pressure force,      
%\begin{equation} 
%\Delta P= - \gamma \nabla\cdot\hat{n}. 
%\label{eq:pressure_dif} 
%\end{equation}
%\noindent This expression is also know as the Young-Laplace equation, a relation describing the capillary pressure across the interface between two fluids, with $\Delta P$ denoting the pressure difference and $\hat{n}$ is the unit normal vector out of the surface.

@@

The MR expression (Eqn.~\ref{eq:MR}) contains phase saturations, $S_{w}\left(x_{i},t\right)$ and  $S_{nw}\left(x_{i},t\right)$ (Eqns.~\ref{Eqn:CoreyBrooks1}-\ref{Eqn:CoreyBrooks2}), however it is clear that during fluid displacement they change in time and space according to the global mass conservation equation. Assuming that phase viscosities remain invariant throughout the simulated fluid displacement, in this work, with no lack of generality, the MR can be replaced by the viscosity ratio (VR),
\begin{displaymath} 
    \text{VR} = \frc{\mu_{i}}{\mu_{j}},
\end{displaymath}
which will be used in the parametrisation of the numerical simulations conducted in Section~\ref{section:results}. Analysis performed in the following sections will make use of this simplified definition as phase saturation $\left(S_{\alpha}\right)$ is a time- and spatial-dependent field which is calculated along with pressure ($p$) and velocity $\left(\mathbf{u}_{\alpha}\right)$.

\medskip
In the next section, the numerical formulation used to simulate multi-fluid flow in porous media is briefly validated (Section \ref{section:results_initial_model_validation}) against laboratory experiments (qualitative validation). The impact of VR (quantitative validation) and heterogeneity on the onset of interface instability and growth of fingers are numerically investigated in Section \ref{section:results_homo_hete}. Section \ref{section:results_hete_fix_adapt} demonstrates the importance of appropriate mesh resolution to adequately capture the initial stages of viscous fingers formation and development. Finally, flow pathway (channelling) is the focus of Section \ref{section:results_3D}.
